---
title: "Netflix Daily Top 10 - Statistical Analysis"
author: "Nazar MYkhailyshchuk, Maryna Ohinska, Daryna Antoniuk"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```

# Introduction

This document provides a comprehensive statistical analysis of the Netflix Daily Top 10 dataset, including descriptive statistics, measures of central tendency, dispersion, and various visualizations.

# Load Required Libraries

```{r libraries}
# Install packages if needed
packages <- c("tidyverse", "moments", "ggplot2", "gridExtra", "knitr", "kableExtra", "psych")

for (pkg in packages) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}
```

# Load and Explore Data

```{r load_data}
# Read the dataset
netflix_data <- read.csv("netflix_daily_top_10.csv", stringsAsFactors = FALSE)

# Display structure
str(netflix_data)

# Display first few rows
head(netflix_data) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Summary statistics
summary(netflix_data)
```

# Data Cleaning and Preparation

```{r data_prep}
# Convert date column to Date type
netflix_data$As.of <- as.Date(netflix_data$As.of)

# Replace "-" with NA in numeric columns and convert to numeric
netflix_data$Year.to.Date.Rank <- as.numeric(ifelse(netflix_data$Year.to.Date.Rank == "-", NA, netflix_data$Year.to.Date.Rank))
netflix_data$Last.Week.Rank <- as.numeric(ifelse(netflix_data$Last.Week.Rank == "-", NA, netflix_data$Last.Week.Rank))

# Ensure other numeric columns are numeric
netflix_data$Rank <- as.numeric(netflix_data$Rank)
netflix_data$Days.In.Top.10 <- as.numeric(netflix_data$Days.In.Top.10)
netflix_data$Viewership.Score <- as.numeric(netflix_data$Viewership.Score)

# Check for missing values
missing_values <- colSums(is.na(netflix_data))
print("Missing Values:")
print(missing_values)

# Get numeric columns for analysis
numeric_cols <- c("Rank", "Year.to.Date.Rank", "Days.In.Top.10", "Viewership.Score")
```

# Descriptive Statistics

## Central Tendency and Dispersion

```{r descriptive_stats}
# Function to calculate mode
get_mode <- function(x) {
  x <- na.omit(x)
  if (length(x) == 0) return(NA)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Calculate statistics for each numeric variable
stats_summary <- data.frame(
  Variable = character(),
  Mean = numeric(),
  Median = numeric(),
  Mode = numeric(),
  SD = numeric(),
  Variance = numeric(),
  Min = numeric(),
  Max = numeric(),
  Range = numeric(),
  IQR = numeric(),
  Skewness = numeric(),
  Kurtosis = numeric(),
  stringsAsFactors = FALSE
)

for (col in numeric_cols) {
  data_col <- netflix_data[[col]]
  data_col_clean <- na.omit(data_col)
  
  # Skip if all values are NA
  if (length(data_col_clean) == 0) {
    stats_summary <- rbind(stats_summary, data.frame(
      Variable = col,
      Mean = NA,
      Median = NA,
      Mode = NA,
      SD = NA,
      Variance = NA,
      Min = NA,
      Max = NA,
      Range = NA,
      IQR = NA,
      Skewness = NA,
      Kurtosis = NA
    ))
    next
  }
  
  stats_summary <- rbind(stats_summary, data.frame(
    Variable = col,
    Mean = round(mean(data_col_clean, na.rm = TRUE), 3),
    Median = round(median(data_col_clean, na.rm = TRUE), 3),
    Mode = get_mode(data_col_clean),
    SD = round(sd(data_col_clean, na.rm = TRUE), 3),
    Variance = round(var(data_col_clean, na.rm = TRUE), 3),
    Min = min(data_col_clean, na.rm = TRUE),
    Max = max(data_col_clean, na.rm = TRUE),
    Range = round(max(data_col_clean, na.rm = TRUE) - min(data_col_clean, na.rm = TRUE), 3),
    IQR = round(IQR(data_col_clean, na.rm = TRUE), 3),
    Skewness = round(skewness(data_col_clean, na.rm = TRUE), 3),
    Kurtosis = round(kurtosis(data_col_clean, na.rm = TRUE), 3)
  ))
}

stats_summary %>%
  kable(caption = "Comprehensive Descriptive Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  scroll_box(width = "100%")
```

## Interpretation of Statistics

```{r interpretation}
cat("INTERPRETATION OF STATISTICAL MEASURES:\n\n")

for (i in 1:nrow(stats_summary)) {
  var_name <- stats_summary$Variable[i]
  skew <- stats_summary$Skewness[i]
  kurt <- stats_summary$Kurtosis[i]
  
  cat(paste0("--- ", var_name, " ---\n"))
  
  # Skewness interpretation
  if (abs(skew) < 0.5) {
    skew_interp <- "approximately symmetric"
  } else if (skew > 0.5) {
    skew_interp <- "positively skewed (right-tailed)"
  } else {
    skew_interp <- "negatively skewed (left-tailed)"
  }
  cat(paste0("Skewness (", skew, "): Distribution is ", skew_interp, "\n"))
  
  # Kurtosis interpretation (excess kurtosis, so 3 is normal)
  if (abs(kurt - 3) < 0.5) {
    kurt_interp <- "mesokurtic (normal peakedness)"
  } else if (kurt > 3) {
    kurt_interp <- "leptokurtic (heavy-tailed, peaked)"
  } else {
    kurt_interp <- "platykurtic (light-tailed, flat)"
  }
  cat(paste0("Kurtosis (", kurt, "): Distribution is ", kurt_interp, "\n\n"))
}
```

# Visualizations

## Viewership Score Analysis

```{r viewership_analysis, fig.height=8}
par(mfrow = c(2, 2))

# Histogram with density curve
hist(netflix_data$Viewership.Score, 
     breaks = 50, 
     main = "Distribution of Viewership Score", 
     xlab = "Viewership Score",
     col = "lightblue",
     border = "white",
     prob = TRUE)
lines(density(netflix_data$Viewership.Score), col = "red", lwd = 2)
abline(v = mean(netflix_data$Viewership.Score), col = "darkgreen", lwd = 2, lty = 2)
abline(v = median(netflix_data$Viewership.Score), col = "purple", lwd = 2, lty = 2)
legend("topright", legend = c("Density", "Mean", "Median"), 
       col = c("red", "darkgreen", "purple"), lwd = 2, lty = c(1, 2, 2))

# Box plot
boxplot(netflix_data$Viewership.Score, 
        main = "Viewership Score Boxplot",
        ylab = "Viewership Score",
        col = "lightgreen",
        border = "darkgreen")

# Q-Q plot for normality
qqnorm(netflix_data$Viewership.Score, main = "Q-Q Plot - Viewership Score")
qqline(netflix_data$Viewership.Score, col = "red", lwd = 2)

# Violin plot using ggplot2
library(ggplot2)
ggplot(netflix_data, aes(x = "", y = Viewership.Score)) +
  geom_violin(fill = "coral", alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Violin Plot - Viewership Score", 
       y = "Viewership Score", x = "")
```

## Days in Top 10 Analysis

```{r days_analysis, fig.height=8}
par(mfrow = c(2, 2))

# Histogram
hist(netflix_data$Days.In.Top.10, 
     breaks = 50, 
     main = "Distribution of Days in Top 10", 
     xlab = "Days in Top 10",
     col = "lightcoral",
     border = "white",
     prob = TRUE)
lines(density(netflix_data$Days.In.Top.10), col = "darkred", lwd = 2)
abline(v = mean(netflix_data$Days.In.Top.10), col = "blue", lwd = 2, lty = 2)
abline(v = median(netflix_data$Days.In.Top.10), col = "green", lwd = 2, lty = 2)
legend("topright", legend = c("Density", "Mean", "Median"), 
       col = c("darkred", "blue", "green"), lwd = 2, lty = c(1, 2, 2))

# Box plot
boxplot(netflix_data$Days.In.Top.10, 
        main = "Days in Top 10 Boxplot",
        ylab = "Days",
        col = "lightyellow",
        border = "orange")

# Q-Q plot
qqnorm(netflix_data$Days.In.Top.10, main = "Q-Q Plot - Days in Top 10")
qqline(netflix_data$Days.In.Top.10, col = "red", lwd = 2)

# Log transformation (since data is skewed)
hist(log(netflix_data$Days.In.Top.10 + 1), 
     breaks = 30, 
     main = "Log-transformed Days in Top 10", 
     xlab = "Log(Days + 1)",
     col = "lavender",
     border = "white")
```

## Rank Distribution

```{r rank_analysis}
# Bar plot of rank frequency
rank_freq <- table(netflix_data$Rank)
barplot(rank_freq, 
        main = "Frequency Distribution of Ranks",
        xlab = "Rank",
        ylab = "Frequency",
        col = rainbow(10),
        border = "white")
```

## Comparison by Content Type

```{r type_comparison, fig.height=8}
# Viewership by Type
ggplot(netflix_data, aes(x = Type, y = Viewership.Score, fill = Type)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Viewership Score by Content Type",
       x = "Type",
       y = "Viewership Score") +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "bottom")

# Days in Top 10 by Type
ggplot(netflix_data, aes(x = Type, y = Days.In.Top.10, fill = Type)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Days in Top 10 by Content Type",
       x = "Type",
       y = "Days in Top 10") +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "bottom")

# Statistical comparison
type_stats <- netflix_data %>%
  group_by(Type) %>%
  summarise(
    Count = n(),
    Mean_Viewership = round(mean(Viewership.Score, na.rm = TRUE), 2),
    Median_Viewership = round(median(Viewership.Score, na.rm = TRUE), 2),
    SD_Viewership = round(sd(Viewership.Score, na.rm = TRUE), 2),
    Mean_Days = round(mean(Days.In.Top.10, na.rm = TRUE), 2),
    Median_Days = round(median(Days.In.Top.10, na.rm = TRUE), 2),
    SD_Days = round(sd(Days.In.Top.10, na.rm = TRUE), 2)
  )

type_stats %>%
  kable(caption = "Statistics by Content Type") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Netflix Exclusive vs Non-Exclusive

```{r exclusive_comparison, fig.height=6}
# Viewership comparison
ggplot(netflix_data, aes(x = Netflix.Exclusive, y = Viewership.Score, fill = Netflix.Exclusive)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Viewership Score: Netflix Exclusive vs Non-Exclusive",
       x = "Netflix Exclusive",
       y = "Viewership Score") +
  scale_fill_manual(values = c("Yes" = "#E50914", "No" = "#221F1F")) +
  theme(legend.position = "bottom")

# Statistical comparison
exclusive_stats <- netflix_data %>%
  group_by(Netflix.Exclusive) %>%
  summarise(
    Count = n(),
    Mean_Viewership = round(mean(Viewership.Score, na.rm = TRUE), 2),
    Median_Viewership = round(median(Viewership.Score, na.rm = TRUE), 2),
    Mean_Days = round(mean(Days.In.Top.10, na.rm = TRUE), 2),
    Median_Days = round(median(Days.In.Top.10, na.rm = TRUE), 2)
  )

exclusive_stats %>%
  kable(caption = "Statistics by Netflix Exclusive Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Time Series Analysis

```{r time_series, fig.height=6}
# Viewership over time
daily_viewership <- netflix_data %>%
  group_by(As.of) %>%
  summarise(Total_Viewership = sum(Viewership.Score, na.rm = TRUE))

ggplot(daily_viewership, aes(x = As.of, y = Total_Viewership)) +
  geom_line(color = "#E50914", size = 1) +
  geom_smooth(method = "loess", color = "blue", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Total Daily Viewership Score Over Time",
       x = "Date",
       y = "Total Viewership Score") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Correlation Analysis

```{r correlation, fig.height=6}
# Select numeric columns for correlation
numeric_data <- netflix_data %>%
  select(Rank, Year.to.Date.Rank, Days.In.Top.10, Viewership.Score) %>%
  na.omit()

# Correlation matrix
cor_matrix <- cor(numeric_data)
print("Correlation Matrix:")
print(round(cor_matrix, 3))

# Visualize correlation matrix
# Install reshape2 if needed
if (!require("reshape2", character.only = TRUE)) {
  install.packages("reshape2", dependencies = TRUE)
  library(reshape2)
}

melted_cor <- melt(cor_matrix)

ggplot(melted_cor, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab",
                       name = "Correlation") +
  theme_minimal() +
  labs(title = "Correlation Heatmap") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
```

## Advanced Visualizations

```{r advanced_viz, fig.height=8}
# Scatter plot with regression line
ggplot(netflix_data, aes(x = Days.In.Top.10, y = Viewership.Score)) +
  geom_point(alpha = 0.3, color = "#E50914") +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  theme_minimal() +
  labs(title = "Relationship between Days in Top 10 and Viewership Score",
       x = "Days in Top 10",
       y = "Viewership Score")

# Density plot by Type
ggplot(netflix_data, aes(x = Viewership.Score, fill = Type)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Viewership Score Distribution by Content Type",
       x = "Viewership Score",
       y = "Density") +
  scale_fill_brewer(palette = "Set1")
```

# Statistical Tests

## Normality Tests

```{r normality_tests}
# Shapiro-Wilk test (sample if dataset is large)
sample_size <- min(5000, nrow(netflix_data))
sample_data <- netflix_data[sample(nrow(netflix_data), sample_size), ]

shapiro_viewership <- shapiro.test(sample_data$Viewership.Score)
shapiro_days <- shapiro.test(sample_data$Days.In.Top.10)

cat("Shapiro-Wilk Normality Test Results:\n\n")
cat("Viewership Score:\n")
cat(paste("  W-statistic:", round(shapiro_viewership$statistic, 4), "\n"))
cat(paste("  p-value:", format(shapiro_viewership$p.value, scientific = TRUE), "\n"))
cat(paste("  Conclusion:", ifelse(shapiro_viewership$p.value < 0.05, 
                                    "Data is NOT normally distributed", 
                                    "Data is normally distributed"), "\n\n"))

cat("Days in Top 10:\n")
cat(paste("  W-statistic:", round(shapiro_days$statistic, 4), "\n"))
cat(paste("  p-value:", format(shapiro_days$p.value, scientific = TRUE), "\n"))
cat(paste("  Conclusion:", ifelse(shapiro_days$p.value < 0.05, 
                                    "Data is NOT normally distributed", 
                                    "Data is normally distributed"), "\n"))
```

## Comparison Tests

```{r comparison_tests}
# T-test: Movies vs TV Shows (Viewership)
movies <- netflix_data$Viewership.Score[netflix_data$Type == "Movie"]
tv_shows <- netflix_data$Viewership.Score[netflix_data$Type == "TV Show"]

t_test_result <- t.test(movies, tv_shows)

cat("Independent T-Test: Movies vs TV Shows (Viewership Score)\n\n")
cat(paste("  t-statistic:", round(t_test_result$statistic, 4), "\n"))
cat(paste("  p-value:", format(t_test_result$p.value, scientific = TRUE), "\n"))
cat(paste("  Conclusion:", ifelse(t_test_result$p.value < 0.05, 
                                    "Significant difference exists", 
                                    "No significant difference"), "\n\n"))

# Wilcoxon test (non-parametric alternative)
wilcox_result <- wilcox.test(movies, tv_shows)
cat("Wilcoxon Rank-Sum Test:\n")
cat(paste("  W-statistic:", wilcox_result$statistic, "\n"))
cat(paste("  p-value:", format(wilcox_result$p.value, scientific = TRUE), "\n"))
```

# Summary and Conclusions

```{r summary}
cat("KEY FINDINGS:\n\n")
cat("1. CENTRAL TENDENCY:\n")
cat(paste("   - Average Viewership Score:", round(mean(netflix_data$Viewership.Score), 2), "\n"))
cat(paste("   - Average Days in Top 10:", round(mean(netflix_data$Days.In.Top.10), 2), "\n\n"))

cat("2. DISTRIBUTION CHARACTERISTICS:\n")
cat(paste("   - Viewership Score Skewness:", round(skewness(netflix_data$Viewership.Score), 3), "\n"))
cat(paste("   - Viewership Score Kurtosis:", round(kurtosis(netflix_data$Viewership.Score), 3), "\n\n"))

cat("3. CONTENT TYPE INSIGHTS:\n")
cat(paste("   - Total Movies:", sum(netflix_data$Type == "Movie"), "\n"))
cat(paste("   - Total TV Shows:", sum(netflix_data$Type == "TV Show"), "\n\n"))

cat("4. VARIABILITY:\n")
cat(paste("   - Viewership Score Standard Deviation:", round(sd(netflix_data$Viewership.Score), 2), "\n"))
cat(paste("   - Days in Top 10 Standard Deviation:", round(sd(netflix_data$Days.In.Top.10), 2), "\n"))
```

---

**Dataset:** Netflix Daily Top 10

**Total Observations:** `r nrow(netflix_data)`
